---
title: "Introduction to linear regression"
author: "Emmanuel Olimi Kasigazi"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

The Human Freedom Index is a report that attempts to summarize the idea
of "freedom" through a bunch of different variables for many countries
around the globe. It serves as a rough objective measure for the
relationships between the different types of freedom - whether it's
political, religious, economical or personal freedom - and other social
and economic circumstances. The Human Freedom Index is an annually
co-published report by the Cato Institute, the Fraser Institute, and the
Liberales Institut at the Friedrich Naumann Foundation for Freedom.

In this lab, you'll be analyzing data from Human Freedom Index reports
from 2008-2016. Your aim will be to summarize a few of the relationships
within the data both graphically and numerically in order to find which
variables can help tell a story about freedom.

## Getting Started

### Load packages

In this lab, you will explore and visualize the data using the
**tidyverse** suite of packages. The data can be found in the companion
package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
data('hfi', package='openintro')
```

### The data

The data we're working with is in the openintro package and it's called
`hfi`, short for Human Freedom Index.

1.  What are the dimensions of the dataset?

```{r}
dim(hfi)
```

```{r}
# Shows dimensions in the console and displays first few rows
head(hfi)
```

```{r}
View(hfi)
```

**[1] 1458 123**

2.  What type of plot would you use to display the relationship between
    the personal freedom score, `pf_score`, and one of the other
    numerical variables? Plot this relationship using the variable
    `pf_expression_control` as the predictor. Does the relationship look
    linear? If you knew a country's `pf_expression_control`, or its
    score out of 10, with 0 being the most, of political pressures and
    controls on media content, would you be comfortable using a linear
    model to predict the personal freedom score?

    For displaying the relationship between the personal freedom score
    (pf_score) and pf_expression_control, a scatterplot would be most
    appropriate since you're examining the relationship between two
    numerical variables.

    ```{r}
    # Creating a scatterplot with pf_expression_control as predictor (x-axis) and pf_score as response (y-axis)
    ggplot(data = hfi, aes(x = pf_expression_control, y = pf_score)) +
      geom_point(alpha = 0.7) +  # Add points with slight transparency for better visualization of overlapping points
      geom_smooth(method = "lm", se = TRUE) +  # Add linear regression line with confidence interval
      labs(
        title = "Relationship between Expression Control and Personal Freedom Score",
        x = "Expression Control Score",
        y = "Personal Freedom Score"
      ) +
      theme_minimal()
    ```

**Looking at the scatterplot, the relationship between
`pf_expression_control` (on the x-axis) and personal freedom score
(`pf_score`, on the y-axis) appears to be quite linear. The blue trend
line shows a positive linear association between the two variables - as
expression control score increases (meaning less political pressure and
control on media), the personal freedom score tends to increase as
well.**

**Key observations: - There's a clear positive correlation between the
variables - The data points follow the linear trend line fairly
consistently across the range of values - The relationship appears to be
reasonably strong, with points clustering around the line - There is
some variance (scatter) around the trend line, but it's relatively
consistent**

**Yes, I would be comfortable using a linear model to predict the
personal freedom score based on a country's `pf_expression_control`
score. The scatter of points shows a pattern that is appropriately
modeled by a straight line, suggesting that a linear model would capture
the relationship well.**

**There is, of course, some variance that wouldn't be explained by this
single predictor, as shown by the vertical spread of points at each
value of `pf_expression_control`, but the linear trend is strong enough
to make reasonably good predictions. A linear regression model would be
suitable for this relationship.**

If the relationship looks linear, we can quantify the strength of the
relationship with the correlation coefficient.

```{r cor}
hfi %>%
  summarise(cor(pf_expression_control, pf_score, use = "complete.obs"))
```

Here, we set the `use` argument to "complete.obs" since there are some
observations of NA.

## Sum of squared residuals

::: {#boxedtext}
In this section, you will use an interactive function to investigate
what we mean by "sum of squared residuals". You will need to run this
function in your console, not in your markdown document. Running the
function also requires that the `hfi` dataset is loaded in your
environment.
:::

Think back to the way that we described the distribution of a single
variable. Recall that we discussed characteristics such as center,
spread, and shape. It's also useful to be able to describe the
relationship of two numerical variables, such as `pf_expression_control`
and `pf_score` above.

3.  Looking at your plot from the previous exercise, describe the
    relationship between these two variables. Make sure to discuss the
    form, direction, and strength of the relationship as well as any
    unusual observations. Form The relationship appears clearly linear.
    The points follow a straight-line pattern from the bottom left to
    the upper right of the plot, and the blue trend line confirms this
    linear pattern. There's no evidence of curvature or a non-linear
    relationship.

**Direction The relationship is positive. As pf_expression_control
increases (meaning less political pressure and control on media
content), the personal freedom score (pf_score) also increases.
Countries with greater freedom of expression tend to have higher overall
personal freedom scores.**

**Strength The relationship is strong, which is confirmed by the
correlation coefficient of 0.796. The points cluster fairly tightly
around the trend line, especially in the middle ranges. The high
correlation indicates that expression control is a substantial component
of and strong predictor for overall personal freedom.**

**Unusual Observations There are a few notable unusual observations:**

**Several outliers appear in the bottom left quadrant where both scores
are very low (countries with highly restricted expression and low
personal freedom) There appears to be one point around expression
control score 7.5 that has an unusually low personal freedom score
(around 5) compared to others with similar expression control scores A
few countries with high expression control scores (7.5-10) show more
variability in their personal freedom scores than might be expected**

**The scatter of points also appears to be somewhat wider at the lower
end of the expression control scale, suggesting that countries with more
media control show more variability in their overall personal freedom
scores than those with freer media.**

Just as you've used the mean and standard deviation to summarize a
single variable, you can summarize the relationship between these two
variables by finding the line that best follows their association. Use
the following interactive function to select the line that you think
does the best job of going through the cloud of points.

```{r plotss-expression-score, eval=FALSE}
# This will only work interactively (i.e. will not show in the knitted document)
hfi <- hfi %>% filter(complete.cases(pf_expression_control, pf_score))
DATA606::plot_ss(x = hfi$pf_expression_control, y = hfi$pf_score)
```

After running this command, you'll be prompted to click two points on
the plot to define a line. Once you've done that, the line you specified
will be shown in black and the residuals in blue. Note that there are 30
residuals, one for each of the 30 observations. Recall that the
residuals are the difference between the observed values and the values
predicted by the line:

$$ e_i = y_i - \hat{y}_i $$

The most common way to do linear regression is to select the line that
minimizes the sum of squared residuals. To visualize the squared
residuals, you can rerun the plot command and add the argument
`showSquares = TRUE`.

```{r plotss-expression-score-squares, eval=FALSE}
DATA606::plot_ss(x = hfi$pf_expression_control, y = hfi$pf_score, showSquares = TRUE)
```

Note that the output from the `plot_ss` function provides you with the
slope and intercept of your line as well as the sum of squares.

4.  Using `plot_ss`, choose a line that does a good job of minimizing
    the sum of squares. Run the function several times. What was the
    smallest sum of squares that you got? How does it compare to your
    neighbors?

**Sum of Squares: 965.674**

## The linear model

It is rather cumbersome to try to get the correct least squares line,
i.e. the line that minimizes the sum of squared residuals, through trial
and error. Instead, you can use the `lm` function in R to fit the linear
model (a.k.a. regression line).

```{r m1}
m1 <- lm(pf_score ~ pf_expression_control, data = hfi)
```

The first argument in the function `lm` is a formula that takes the form
`y ~ x`. Here it can be read that we want to make a linear model of
`pf_score` as a function of `pf_expression_control`. The second argument
specifies that R should look in the `hfi` data frame to find the two
variables.

The output of `lm` is an object that contains all of the information we
need about the linear model that was just fit. We can access this
information using the summary function.

```{r summary-m1}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to
describe the model is shown at the top. After the formula you find the
five-number summary of the residuals. The "Coefficients" table shown
next is key; its first column displays the linear model's y-intercept
and the coefficient of `pf_expression_control`. With this table, we can
write down the least squares regression line for the linear model:

$$ \hat{y} = 4.61707 + 0.49143 \times pf\_expression\_control $$

One last piece of information we will discuss from the summary output is
the Multiple R-squared, or more simply, $R^2$. The $R^2$ value
represents the proportion of variability in the response variable that
is explained by the explanatory variable. For this model, 63.42% of the
variability in runs is explained by at-bats.

5.  Fit a new model that uses `pf_expression_control` to predict
    `hf_score`, or the total human freedom score. Using the estimates
    from the R output, write the equation of the regression line. What
    does the slope tell us in the context of the relationship between
    human freedom and the amount of political pressure on media content?

```{r m2}
m2 <- lm(hf_score ~ pf_expression_control, data = hfi)
```

```{r}
summary(m2)
```

**hf_score = 5.153687 + 0.349862 × pf_expression_control Interpreting
the slope (0.349862) in context:**

**For each one-unit increase in the expression control score (meaning
less political pressure and control on media content), we expect the
total human freedom score to increase by approximately 0.35 points,
holding all else constant.**

**This tells us that there's a significant positive relationship between
media freedom and overall human freedom. Countries where there is less
political pressure on media content tend to have higher total human
freedom scores.**

**The relationship is strong and statistically significant (p \<
2e-16).**

**The R-squared value of 0.5775 indicates that approximately 58% of the
variation in total human freedom scores can be explained by the level of
expression control alone.**

**This is slightly lower than what we saw for personal freedom (63.4%),
which makes sense because total human freedom also incorporates economic
freedom components that might be less directly connected to media
freedom.**

**The positive slope confirms that societies with freer media
environments tend to have greater overall human freedom, including both
personal and economic dimensions.** **Insert your answer here**

## Prediction and prediction errors

Let's create a scatterplot with the least squares line for `m1` laid on
top.

```{r reg-with-line}
ggplot(data = hfi, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```

Here, we are literally adding a layer on top of our plot. `geom_smooth`
creates the line by fitting a linear model. It can also show us the
standard error `se` associated with our line, but we'll suppress that
for now.

This line can be used to predict $y$ at any value of $x$. When
predictions are made for values of $x$ that are beyond the range of the
observed data, it is referred to as *extrapolation* and is not usually
recommended. However, predictions made within the range of the data are
more reliable. They're also used to compute the residuals.

6.  If someone saw the least squares regression line and not the actual
    data, how would they predict a country's personal freedom school for
    one with a 6.7 rating for `pf_expression_control`? Is this an
    overestimate or an underestimate, and by how much? In other words,
    what is the residual for this prediction?

**For a country with pf_expression_control = 6.7: Predicted pf_score =
4.61707 + 0.49143 × 6.7 Predicted pf_score = 4.61707 + 3.29258 Predicted
pf_score = 7.90965 Therefore, if someone only saw the regression line
and not the actual data, they would predict a personal freedom score of
approximately 7.91 for a country with an expression control rating of
6.7. To determine if this is an overestimate or underestimate, I would
need to know the actual pf_score for a country with
pf_expression_control = 6.7. I looked at the data adn saw several
countries with 6.75 but no 6.7 pf_expression_control.**

**Looking at the scatterplot, I can see that at x = 6.7, there are
multiple data points with varying y-values both above and below the
regression line.**

**For countries at this expression control level, some have personal
freedom scores higher than 7.91 (the prediction would be an
underestimate for these), while others have lower scores (the prediction
would be an overestimate for these).**

**The residual would be calculated as: Residual = Actual pf_score -
Predicted pf_score = Actual pf_score - 7.9**

## Model diagnostics

To assess whether the linear model is reliable, we need to check for (1)
linearity, (2) nearly normal residuals, and (3) constant variability.

**Linearity**: You already checked if the relationship between
`pf_score` and \`pf_expression_control' is linear using a scatterplot.
We should also verify this condition with a plot of the residuals vs.
fitted (predicted) values.

```{r residuals}
ggplot(data = m1, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```

Notice here that `m1` can also serve as a data set because stored within
it are the fitted values ($\hat{y}$) and the residuals. Also note that
we're getting fancy with the code here. After creating the scatterplot
on the first layer (first line of code), we overlay a horizontal dashed
line at $y = 0$ (to help us check whether residuals are distributed
around 0), and we also reanme the axis labels to be more informative.

7.  Is there any apparent pattern in the residuals plot? What does this
    indicate about the linearity of the relationship between the two
    variables?

    **Looking at the residuals vs. fitted values plot, I don't see any
    strong or systematic pattern in the residuals. The points appear to
    be randomly scattered around the horizontal dashed line at y=0, with
    no clear trends, curves, or funnels visible.**

**Key observations about this residuals plot: 1. The residuals are
fairly evenly distributed above and below zero across all fitted values
2. There's no obvious curvature or trend that would suggest nonlinearity
3. The spread of residuals appears relatively consistent across the
range of fitted values**

**This lack of pattern in the residuals plot indicates that the linear
model is appropriate for describing the relationship between personal
freedom score (pf_score) and expression control (pf_expression_control).
The random scatter of residuals supports the linearity assumption in our
regression model.**

**If there had been a curved pattern in the residuals, it would have
suggested that a linear model was not capturing the true relationship
between these variables, and a nonlinear model might be more
appropriate. But based on this plot, the linear model appears to be a
suitable fit for the data.**

**Nearly normal residuals**: To check this condition, we can look at a
histogram

```{r hist-res}
ggplot(data = m1, aes(x = .resid)) +
  geom_histogram(binwidth = 0.5) +
  xlab("Residuals")
```

or a normal probability plot of the residuals.

```{r qq-res}
ggplot(data = m1, aes(sample = .resid)) +
  stat_qq()
```

Note that the syntax for making a normal probability plot is a bit
different than what you're used to seeing: we set `sample` equal to the
residuals instead of `x`, and we set a statistical method `qq`, which
stands for "quantile-quantile", another name commonly used for normal
probability plots.

8.  Based on the histogram and the normal probability plot, does the
    nearly normal residuals condition appear to be met?

    **Based on both the histogram and the normal probability plot (Q-Q
    plot), the nearly normal residuals condition appears to be
    reasonably well met for this linear model.**

**Looking at the histogram of residuals: - The distribution appears
approximately symmetric and bell-shaped - It's centered around zero -
There's a slight positive skew (the right tail is slightly longer than
the left) - The bulk of residuals fall between -2 and 2**

**The Q-Q plot also supports this conclusion: - The points follow the
theoretical normal line quite well through the middle of the
distribution - There are some minor deviations at the extreme tails,
especially at the very high and very low ends - These small deviations
at the tails are common in real-world data and not severe enough to
invalidate the model**

**While the residuals aren't perfectly normal (which is rarely the case
with real data), they are sufficiently close to normal for the purposes
of linear regression. The slight departures from normality aren't
extreme enough to undermine the validity of the model or its statistical
inferences.**

**Therefore, the nearly normal residuals condition appears to be met for
this linear regression model.**

**Constant variability**:

9.  Based on the residuals vs. fitted plot, does the constant
    variability condition appear to be met?

**In that plot, we're checking whether the spread of residuals remains
consistent across all fitted values. Ideally, the vertical spread of
residuals should be roughly the same regardless of where you look along
the x-axis (fitted values).**

**From the plot , I observe that: - The residuals appear to have a
fairly consistent spread across the range of fitted values - There's no
obvious funnel shape (widening or narrowing of the residuals) - The
cloud of points maintains a similar vertical spread from the left to the
right side of the plot**

**There might be a very slight tendency for the spread to be narrower at
the extreme ends of the fitted values range, but this is not pronounced
and could be due to having fewer observations at those extremes.**

**Overall, the constant variability condition appears to be reasonably
well met for this linear regression model. The relatively uniform spread
of residuals across fitted values suggests that the model's predictions
have similar precision regardless of the value being predicted, which
supports the validity of the model's inferences and predictions.**

------------------------------------------------------------------------

## More Practice

-   Choose another freedom variable and a variable you think would
    strongly correlate with it.. Produce a scatterplot of the two
    variables and fit a linear model. At a glance, does there seem to be
    a linear relationship?

    A good pair to explore would be economic freedom (ef_score) and rule
    of law (pf_rol), as rule of law is often theorized to be a
    foundation for economic freedom.

    ```{r rule of law adn econ freedom}
    # Create a scatterplot of ef_score vs pf_rol
    ggplot(data = hfi, aes(x = pf_rol, y = ef_score)) +
      geom_point(alpha = 0.7) +
      geom_smooth(method = "lm", se = TRUE) +
      labs(
        title = "Relationship between Rule of Law and Economic Freedom",
        x = "Rule of Law Score",
        y = "Economic Freedom Score"
      ) +
      theme_minimal()

    ```

```{r m3 linear}
# Fit a linear model
m3 <- lm(ef_score ~ pf_rol, data = hfi)
summary(m3)

```

```{r correlation check}
# Check correlation
cor(hfi$pf_rol, hfi$ef_score, use = "complete.obs")
```

Looking at the scatterplot of Rule of Law vs. Economic Freedom, there
appears to be a clear positive linear relationship between these
variables. This is confirmed by the correlation coefficient of
approximately 0.71, which indicates a strong positive correlation.

Key observations from the plot:

1.  **Linear Trend**: The blue regression line shows a positive slope,
    indicating that countries with higher rule of law scores tend to
    have higher economic freedom scores.

2.  **Strength of Relationship**: With a correlation of 0.71, about 50%
    (0.71² ≈ 0.5) of the variation in economic freedom can be explained
    by rule of law.

3.  **Data Distribution**: The points follow the linear trend reasonably
    well, though there is some scatter around the line, indicating that
    other factors also influence economic freedom.

4.  **Interesting Features**:

    -   There appears to be more variability in economic freedom scores
        at the middle range of rule of law scores
    -   At very high rule of law scores (7.5+), there seems to be a
        cluster of countries with very high economic freedom (8+)
    -   There are a few outliers, including some countries with
        relatively low rule of law but moderate economic freedom

This relationship makes theoretical sense: strong rule of law provides
the legal framework and stability necessary for economic freedom to
flourish. Countries with clear, consistent, and fairly enforced laws
tend to have better property rights protection, contract enforcement,
and regulatory environments - all important components of economic
freedom.

The linear relationship appears appropriate for modeling this data, and
a linear regression model would be suitable for predicting economic
freedom based on rule of law.

Let's Also explore the relationship between rule of law in criminal
justice (pf_rol_criminal) and personal freedom score (pf_score). The
criminal justice system's fairness and effectiveness is an important
component of personal freedom, and it would be interesting to see how
strongly it correlates with overall personal freedom.

```{r}
# Create a scatterplot of pf_score vs pf_rol_criminal
ggplot(data = hfi, aes(x = pf_rol_criminal, y = pf_score)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Relationship between Criminal Justice and Personal Freedom",
    x = "Rule of Law - Criminal Justice Score",
    y = "Personal Freedom Score"
  ) +
  theme_minimal()

# Fit a linear model
m4 <- lm(pf_score ~ pf_rol_criminal, data = hfi)
summary(m4)

# Check correlation
cor(hfi$pf_rol_criminal, hfi$pf_score, use = "complete.obs")
```

Looking at the scatterplot of Rule of Law - Criminal Justice Score vs.
Personal Freedom Score, there's a clear positive linear relationship
between these variables. The correlation coefficient of approximately
0.68 confirms this is a strong positive correlation.

Key observations from this analysis:

1.  **Linear Relationship**: The blue regression line shows a clear
    positive slope, indicating that countries with better criminal
    justice systems tend to have higher personal freedom scores.

2.  **Strength of Relationship**: With a correlation of 0.68, about
    46.6% (0.68² ≈ 0.466) of the variation in personal freedom scores
    can be explained by the criminal justice component of rule of law.

3.  **Comparative Analysis**:

    -   This relationship (r = 0.68, R² ≈ 0.47) is stronger than the
        relationship between rule of law and economic freedom (r = 0.71,
        R² ≈ 0.50)
    -   However, it's slightly weaker than the relationship between
        expression control and personal freedom (r = 0.796, R² ≈ 0.63)

4.  **Data Distribution**:

    -   There's more variability in the personal freedom scores at the
        middle range of criminal justice scores
    -   The relationship appears most consistent at the higher end of
        the scale
    -   There are some outliers, particularly countries with moderately
        high criminal justice scores but lower personal freedom

This relationship makes intuitive sense: fair criminal justice systems
that respect due process and human rights are foundational to personal
freedom. Countries where individuals are protected from arbitrary
detention, have fair trials, and face proportionate punishments tend to
have higher overall personal freedom.

The strength of this relationship suggests that criminal justice is an
important component of personal freedom, though not as predictive as
freedom of expression. This could be because expression rights might
have more direct day-to-day impacts on citizens' sense of freedom than
criminal justice, which affects fewer people directly but remains
essential for the rule of law.

**Insert your answer here**

-   How does this relationship compare to the relationship between
    `pf_expression_control` and `pf_score`? Use the $R^2$ values from
    the two model summaries to compare. Does your independent variable
    seem to predict your dependent one better? Why or why not?

Rule of Law vs. Economic Freedom To compare these relationships, I need
to look at the R² values from both models:

1.  For the relationship between `pf_expression_control` and `pf_score`:
    -   R² = 0.6342 (from the earlier model summary)
2.  For the relationship between `pf_rol` (rule of law) and `ef_score`
    (economic freedom):
    -   We can calculate this from the correlation: R² = 0.7064² ≈ 0.499

Comparing these R² values: - The expression control model explains about
63.4% of the variation in personal freedom scores - The rule of law
model explains about 49.9% of the variation in economic freedom scores

This means that `pf_expression_control` appears to be a better predictor
of `pf_score` than `pf_rol` is of `ef_score`, though both show strong
positive relationships.

Why might this be the case?

1.  **Component vs. Composite Relationship**: Expression control is one
    of the components that directly makes up the personal freedom score,
    so it's expected to have a stronger relationship. Rule of law, while
    important for economic freedom, might have a more indirect
    relationship with the overall economic freedom score.

2.  **Multiple Dimensions of Economic Freedom**: Economic freedom likely
    depends on many diverse factors besides rule of law, such as tax
    policies, trade regulations, and monetary stability. Personal
    freedom might have a more cohesive structure with expression being a
    core element.

3.  **Measurement Precision**: The way these variables are measured
    could affect how strongly they correlate. If personal freedom
    metrics are more internally consistent than economic freedom
    metrics, we would expect stronger correlations among the former.

Both relationships show important connections between foundational
aspects of freedom and overall freedom scores, but the data suggests
that expression control is more strongly predictive of overall personal
freedom than rule of law is of economic freedom.

Rule of Law - Criminal Justice Score vs. Personal Freedom Comparing the
relationships between these variables using R² values:

1.  For `pf_expression_control` and `pf_score`:
    -   Correlation = 0.796
    -   R² = 0.634 (63.4% of variation explained)
2.  For `pf_rol_criminal` and `pf_score`:
    -   Correlation = 0.682
    -   R² = 0.466 (46.6% of variation explained)

Based on these values, `pf_expression_control` is a stronger predictor
of overall personal freedom than criminal justice systems. The
expression control variable explains approximately 17 percentage points
more variation in personal freedom scores than the criminal justice
variable.

Why might this be the case?

1.  **Direct vs. Indirect Impact**: Freedom of expression might have a
    more direct and widespread impact on daily life for the average
    citizen. While criminal justice is crucial, many people might not
    directly interact with the criminal justice system, whereas
    expression freedoms affect nearly everyone daily.

2.  **Composite Score Weighting**: The personal freedom score likely
    includes expression freedom as a more heavily weighted component
    than criminal justice metrics, which would naturally lead to a
    stronger correlation.

3.  **Leading Indicator**: Expression freedom often serves as a canary
    in the coal mine for overall personal freedom. When governments
    begin restricting freedoms, expression is frequently targeted first.

4.  **Measurement Clarity**: Freedom of expression might be easier to
    measure consistently across countries than criminal justice
    fairness, which involves more complex and varied systems.

Both variables are strong predictors of personal freedom, which makes
sense conceptually. However, the data indicates that knowing a country's
level of media freedom and expression control gives us more predictive
power about its overall personal freedom than knowing the quality of its
criminal justice system.

This illustrates how different components of freedom vary in their
relationship to overall freedom measures, providing insight into the
complex structure of what makes societies free. **Insert your answer
here**

-   What's one freedom relationship you were most surprised about and
    why? Display the model diagnostics for the regression model
    analyzing this relationship.

    Let me explore a freedom relationship I find particularly surprising
    by examining the relationship between religious freedom and economic
    freedom.

```{r}
# Create a scatterplot of ef_score vs pf_religion
ggplot(data = hfi, aes(x = pf_religion, y = ef_score)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Relationship between Religious Freedom and Economic Freedom",
    x = "Religious Freedom Score",
    y = "Economic Freedom Score"
  ) +
  theme_minimal()

```

I find this relationship surprising because economic freedom and
religious freedom come from different conceptual traditions. Religious
freedom is primarily about personal beliefs and worship practices, while
economic freedom concerns markets, property rights, and commercial
activities. While we might expect democratic societies to have both,
there are notable examples of countries with significant religious
restrictions yet relatively free economies (like some Gulf states), and
others with strong.

What surprises me about this relationship is how it reveals the
interconnectedness of different types of freedom, despite their
conceptual differences. It suggests that societies might approach
freedom more holistically than we sometimes assume, rather than treating
different freedoms as entirely separate domains.religious freedoms but
more controlled economies.

```{r}
# Fit a linear model
m5 <- lm(ef_score ~ pf_religion, data = hfi)
summary(m5)

# Check correlation
cor(hfi$pf_religion, hfi$ef_score, use = "complete.obs")
```

```{r}
# Model diagnostics
# 1. Residuals vs Fitted plot
ggplot(data = m5, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted values", 
       y = "Residuals")
```

```{r}
# 2. Q-Q plot for normality of residuals
ggplot(data = m5, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles", 
       y = "Sample Quantiles")
```

```{r}
# 3. Histogram of residuals
ggplot(data = m5, aes(x = .resid)) +
  geom_histogram(binwidth = 0.5) +
  labs(title = "Histogram of Residuals",
       x = "Residuals", 
       y = "Count")
```

Looking at these results, I'm genuinely surprised by the relationship
between religious freedom and economic freedom. This is perhaps the most
unexpected finding in our analysis of the Human Freedom Index data.

The correlation coefficient is only 0.132, which indicates a very weak
positive relationship between religious freedom and economic freedom.
This translates to an R² of only about 0.017, meaning that religious
freedom explains less than 2% of the variation in economic freedom
scores.

What makes this relationship surprising:

1.  **Weak Association**: Unlike the other freedom relationships we
    examined, which showed moderate to strong correlations (0.68-0.80),
    religious freedom and economic freedom are barely related. The
    nearly flat regression line in the scatterplot confirms this visual
    impression.

2.  **Theoretical Implications**: This challenges the assumption that
    different types of freedom are strongly interconnected. While
    personal freedoms like expression and criminal justice protections
    showed strong relationships, religious freedom appears to operate
    more independently from economic systems.

3.  **Real-World Examples**: This data helps explain why we see
    countries with high religious freedom but varying levels of economic
    freedom (e.g., some European welfare states), and others with high
    economic freedom but restricted religious practices (e.g., certain
    East Asian and Middle Eastern economies).

The weak relationship between religious freedom and economic freedom
does make intuitive sense from several perspectives, and I don't think
our data is wrong. Here's why:

1.  **Different Historical Developments**: Religious freedom and
    economic freedom often developed through separate historical
    processes. Religious freedom emerged through religious wars,
    enlightenment thinking, and constitutional protections, while
    economic freedom developed through commercial revolutions,
    industrialization, and economic liberalization policies.

2.  **Real-World Examples Support This Finding**:

    -   Singapore has high economic freedom but moderate religious
        freedom with some restrictions
    -   Some European countries (like Sweden or Denmark) have very high
        religious freedom but more regulated economies
    -   Gulf states like UAE or Qatar have relatively high economic
        freedom scores but significant religious restrictions
    -   India has strong religious freedom protections constitutionally
        but more moderate economic freedom scores

3.  **Different Institutional Requirements**: Religious freedom
    primarily requires non-intervention by the state in personal beliefs
    and practices, while economic freedom requires specific
    institutional arrangements like property rights enforcement,
    contract law, and particular regulatory environments.

4.  **Cultural vs. Structural Factors**: Religious freedom is often more
    culturally determined, while economic freedom is more structurally
    determined through formal institutions and policies.

The HFI is a well-respected dataset compiled by multiple research
institutes (Cato Institute, Fraser Institute, and Liberales Institut),
so it's likely that this weak correlation genuinely reflects reality
rather than a data problem.

This finding challenges the notion that all freedoms necessarily move
together, suggesting that human freedom is more multidimensional than
sometimes assumed. Some societies prioritize different types of freedoms
based on their specific cultural, historical, and political contexts.

This complexity actually makes our analysis more interesting - it
reveals that the relationships between different types of freedom aren't
uniform, which is valuable information for understanding how freedom
develops across societies.

The model diagnostics support these observations:

-   **Residuals vs. Fitted Plot**: Shows no clear pattern but
    demonstrates wide scatter, confirming the weak predictive power of
    the model.
-   **Q-Q Plot**: Indicates the residuals follow a reasonably normal
    distribution, though with some deviation at the tails.
-   **Histogram of Residuals**: Shows a roughly normal distribution,
    though slightly skewed.

This finding suggests that religious freedom and economic freedom may
develop through different historical, cultural, and political
mechanisms. A society might prioritize religious tolerance without
embracing free markets, or vice versa. This independence between
religious and economic freedoms stands in stark contrast to the stronger
relationships we observed between other freedom variables.

The weak relationship also suggests that policy approaches targeting one
type of freedom may not automatically enhance other types, requiring
more nuanced and specific approaches to promoting different dimensions
of human freedom. **Insert your answer here**

------------------------------------------------------------------------
